---
title: "The effect of dormancy on extinction probability"
output: html_notebook
---

Based on:
Lennon, J. T., den Hollander, F., Wilke-Berenguer, M. & Blath, J. Principles of seed banks and the emergence of complexity from dormancy. Nature Communications 12, (2021).

Usually only external forces such as environmental conditions or resource availability are considered when looking at fluctuations and competition between species, but the ability to enter dormancy can have a significant impact in how successful a species is during competition. This model will model how extinction rates across phylogeny vary with dormancy.

-Species1 equilibrium reproduction rate = (lambda1 > 0)
-Species2 equilibrium reproduction rate = (lambda2 < lambda1)
    reduced because it reflects the cost of them being able to engage in dormancy
-Species1 equilibrium population size = lambda1 - mu
-Species2 equilibrium population size = (lambda2 - mu)/(1 - p)
-Species1 death rate = (death1 < lambda2)
-Species2 death rate = (death2 < lambda2)
-Species1 active individuals = N1
-Species2 active individuals = N2a
-All active individuals = Na = sum(N1 and N2a)
-Competition is proportional to Na
-Species1 removal = Na
-Species2 escape probability = p
-Species2 removal = (1 - p)Na
-Species2 competition-induced dormancy rate = pNa
-Species2 resuscitation rate = 1

Species2 fixation possible if (lambda1 - mu) < (lamabda2 - mu)/(1 - p)

###########################
-Taxa with a larger population size are less likely to go extinct
-Taxa with dormancy are more likely to have a larger population size
-Taxa with large population sizes are more likely to speciate
-Parameter we care about is number of species going extinct
-Evolution of dormancy informs population growth patterns which affects dormancy


```{r}
S1equilreproductionrate <- lambda1
S2equilreproductionrate <- lambda2
S1equilpopsize <- S1equilreproductionrate - mu
S2equilpopsize <- (S2equilreproductionrate - mu)/(1 - p)
S1deathrate <- death1
S2deathrate <- death2
S1active <- N1
S2active <- N2a
Allactiveind <- Na = (N1 + N2a)
S1removalrate <- Allactiveind
S2escapeprob <- p
S2removal <- (1 - p)Allactiveind
S2dormancyrate <- S2escapeprob * Allactiveind
S2resuscitationrate <- 1

```

Sp1 = Not dormancy capable species
Sp2 = Dormancy capable species
Re = Reproduction rate
De = Death rate
Ex = Extinction rate
Po = Population size
Es = Escape rate
```{r}
Sp2_Re <- 
Sp2s_De <- 
Sp2_Es <- 
Sp2_Po <- 
  
Sp1_Re <- 
Sp1_De <- 
Sp1_Po <- 
  
Sp2_SpeciationRate <- 
Sp2_ExtinctionRate <- 
Sp1_SpeciationRate <- 
Sp1_ExtinctionRate <- 
```

