---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

Following along with Dr. O'Meara's correlation models assignment (http://phylometh.info/correlations.html).
```{r}
library(geiger)
library(ape)
library(corHMM)
# using the data included with the assignment
tree.primates <- read.tree(text = "((((Homo:0.21,Pongo:0.21):0.28,Macaca:0.49):0.13,Ateles:0.62):0.38,Galago:1.00);") # make a tree
X <- c(4.09434, 3.61092, 2.37024, 2.02815, -1.46968) # what are these? Two different traits?
Y <- c(4.74493, 3.33220, 3.36730, 2.89037, 2.30259)

names(X) <- names(Y) <- c("Homo", "Pongo", "Macaca", "Ateles", "Galago") # add names to the data. 
pic.X <- pic(X, tree.primates) # compute the phylogenetically independent contrasts using the method described by Felsenstein (1985)
pic.Y <- pic(Y, tree.primates)
```

Positivitize the contrasts and do a regression through the origin. 
```{r}
require("corHMM") # just like library(), but is for packages that are used within a function.
# corHMM extimates hidden rates underlying the evolution of a binary character. It's a hidden rates model. 
data(primates)
ls() # lists all data and values in the environment
print(primates)
require(phytools)
```

We would like to have all 4 trait combinations present for this exercise, but it's making up data, so it's just for educational purposes!
```{r}
primates$trait[which(grepl("Hylobates", primates$trait[,1])),2] <- 1

trait1 <- primates$trait[,2]
names(trait1) <- primates$trait[,1]
primates$tree <- ape::multi2di(primates$tree)
plotSimmap(make.simmap(primates$tree, trait1), 
           pts = FALSE, 
           fsize = 0.8)
rate.mat.er <- corHMM:::rate.mat.maker(rate.cat = 1,
                              hrm = FALSE,
                              ntraits = 1,
                              nstates = 2,
                              model = "ER") # equal rates model
# makes a rate matrix, can make a matrix without this code if needed.
# rate matrix maker in the corHMM package doesn't exist anymore, so use ::: to find it :)
print(rate.mat.er)
```

This matrix is the probability of changing between the different states. In this example, it is equally probable to change from 1 to 2 as it is from 2 to 1.
```{r}
pp.er <- corHMM(phy = primates$tree,
                data = primates$trait[,c(1,2)],
                rate.cat = 1,
                rate.mat = rate.mat.er,
                node.states = "marginal")
print(pp.er)
```
These are the optimized rates we entered for rate.mat.er. Inferred ancestral states using marginal reconstruction. Estimates hidden rates underlying the evolution of a binary character.
The results mean that the likelihood of switching from 1->2 and 2->1 is still equal.

```{r}
rate.mat.ard <- corHMM:::rate.mat.maker(rate.cat = 1,
                                        hrm = FALSE,
                                        ntraits = 1,
                                        nstates = 2,
                                        model = "ARD") # ARD = all rates different model
print(rate.mat.ard)
```
The matrix has two different rates for the two different switches

```{r}
pp.ard <- corHMM(phy = primates$tree,
                 data = primates$trait[,c(1,2)],
                 rate.cat = 1,
                 rate.mat = rate.mat.ard,
                 node.states = "marginal")
print(pp.ard)
```
These are the optimized rates, but because we used the "All Rates Different" model the rates are different.

Which model is better?
```{r}
print(paste("Equal rates model AICc =", pp.er$AICc, "All rates different model AICc =", pp.ard$AICc))
```
The equal rates model has a lower AICc, so it fits the data better.

Multiple Traits

Matrix with 4 states
```{r}
rate.mat.er.4state <- corHMM:::rate.mat.maker(rate.cat = 1,
                                              hrm = FALSE,
                                              ntraits = 1,
                                              nstates = 4,
                                              model = "ER")
print(rate.mat.er.4state)
```

Now convert the two binary traits into a single four character state.
```{r}
fourstate.trait <- rep(NA, Ntip(primates$tree))
for(i in sequence(Ntip(primates$tree))) {
  if(primates$trait[i,2] == 0 && primates$trait[i,3] == 0) {
    fourstate.trait[i] <- 0
  }
  if(primates$trait[i,2] == 0 && primates$trait[i,3] == 1) {
    fourstate.trait[i] <- 1
  }
  if(primates$trait[i,2] == 1 && primates$trait[i,3] == 0) {
  fourstate.trait[i] <- 2
  }
  if(primates$trait[i,2] == 1 && primates$trait[i,3] == 1) {
  fourstate.trait[i] <- 3
  }
}
fourstate.data <- data.frame(Genus_sp = primates$trait[,1],
                             T1 = fourstate.trait)
print(rayDISC(phy = primates$tree,
              data = fourstate.data,
              ntraits = 1,
              model = "ER",
              node.states = "marginal"))
# rayDISC fits a model of evolution for categorical traits (multistate ok, polymorphisms ok, missing data ok, and incomplete resolution on tree ok).
print(rayDISC(phy = primates$tree,
              data = fourstate.data,
              ntraits = 1,
              rate.mat = rate.mat.er.4state,
              model = "ARD",
              node.states = "marginal"))
rate.mat.ard.4state <- corHMM:::rate.mat.maker(rate.cat = 1,
                                               hrm = FALSE,
                                               ntraits = 1,
                                               nstates = 4,
                                               model = "ARD")
print(rate.mat.ard.4state)
```

GTR (generalized time reversible) matrix:
```{r}
rate.mat.gtr.4states <- rate.mat.ard.4state
rate.mat.gtr.4states <- corHMM:::rate.par.eq(rate.mat.index = rate.mat.gtr.4states, eq.par = c(1,4))
rate.mat.gtr.4states <- corHMM:::rate.par.eq(rate.mat.index = rate.mat.gtr.4states, eq.par = c(2,6))
rate.mat.gtr.4states <- corHMM:::rate.par.eq(rate.mat.index = rate.mat.gtr.4states, eq.par = c(3,8))
rate.mat.gtr.4states <- corHMM:::rate.par.eq(rate.mat.index = rate.mat.gtr.4states, eq.par = c(4,6))
rate.mat.gtr.4states <- corHMM:::rate.par.eq(rate.mat.index = rate.mat.gtr.4states, eq.par = c(5,7))
rate.mat.gtr.4states <- corHMM:::rate.par.eq(rate.mat.index = rate.mat.gtr.4states, eq.par = c(6,7))

print(rate.mat.gtr.4states) # we went through and made the top-right half of the matrix mirror the bottom-left half of the matrix (reversible!)

print(rayDISC(phy = primates$tree,
              data = fourstate.data,
              ntraits = 1,
              rate.mat = rate.mat.gtr.4states,
              node.states = "marginal",
              model = "ARD"))
```
These are the optimized rates for each change (but reversible).

Make a Pagel 1994 model
```{r}
print(corHMM:::rate.mat.maker(rate.cat = 1,
                              hrm = FALSE,
                              ntraits = 2,
                              nstates = 2,
                              model = "ARD"))
rate.mat.pag94 <- corHMM:::rate.par.drop(rate.mat.ard.4state,
                                         drop.par = c(3,5,8,10))
print(rate.mat.pag94)
```

Optimize the rates for Pagel 94.
```{r}
print(rayDISC(phy = primates$tree,
              data = fourstate.data,
              ntraits = 1,
              rate.mat = rate.mat.pag94,
              node.states = "marginal",
              model = "ARD"))
```

Read and Nee (1995) suggested a paired design - similar to a medical study where one person receives a treatment, the other doesn't and that counts as one sample. If you pair the data from one taxa with the other, then it's a similar experimental design. 
Their example of warning coloration and gregariousness says something about having warning color (first trait) increases the chance that it would be gregarious (second trait), so a rate matrix would show a higher rate for 0,0 -> 1,1 than expected based on the rate for just 0,0 -> 1,0. 







