---
title: "Discrete Characters"
date: "3/9/2022"
output: html_document
---

Following along with the Discrete Characters exercise by Dr. O'Meara: http://phylometh.info/discrete.html
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rotl)
library(ape)
library(RCurl)
library(geiger)
library(dispRity)
library(tidyverse)
library(Hmisc)
library(phangorn)
```

Borrowed data from https://lukejharmon.github.io/pcm/rmarkdown/chapter8_squamatelimbs.html
```{r, eval=TRUE}
tree <- read.tree(text=getURL("https://raw.githubusercontent.com/lukejharmon/pcm/master/datafiles/squamate.phy"))
plot.phylo(tree)
```

Character data
```{r, eval=TRUE}
discrete.data <- read.csv(text=getURL("https://raw.githubusercontent.com/lukejharmon/pcm/master/datafiles/brandley_table.csv"))

rownames(discrete.data) <- gsub(" ", "_", discrete.data$Species)

# to simplify the data I just went down to the Morph column.
subset.discrete.data <- discrete.data[, c("Species", "Morph")]

```

Clean the data. Needed to remove 3 taxa after running the function.
```{r, eval=TRUE}
CleanData <- function(phy, data) {
  treedata(tree, 
           subset.discrete.data, 
           sort = FALSE, 
           warnings = TRUE)
}
CleanData(tree, subset.discrete.data)
cleaned.data <- subset.discrete.data[!(subset.discrete.data$Species == "Gonatodes_albogularis" |
                                       subset.discrete.data$Species == "Lepidophyma_flavimaculatum" |
                                       subset.discrete.data$Species == "Trachyboa_boulengeri"),]
cleaned.data <- cleaned.data %>%
  mutate_at(vars(Morph), .funs = as.factor) %>%
  select(-Species)
```

Visualize the discrete data
It's a bit messy looking, but it does show histograms of the data. If you knew what the data was supposed to look like, it might make sense. 
```{r, eval=TRUE}
par(mar=c(1,1,1,1))
VisualizeData <- function(phy, data) {
  hist.data.frame(cleaned.data)
}
VisualizeData(tree, cleaned.data)
```

Visualize the tree. Doesn't seem like a lot of problems with polytomies. 
```{r, eval=TRUE}
plot.phylo(tree)
```

Look at ancestral states with parsimony:
```{r, eval=TRUE}
a <- c("0", "1", "2", "NA")
cleaned.discrete.phyDat <- phyDat(cleaned.data, type = "USER", levels = levels(as.factor(a)))
# kept all the levels for Toes, Fingers, Morph, and Ecology.1

anc.p <- ancestral.pars(tree, cleaned.discrete.phyDat)
plotAnc(tree, data = anc.p, i = 1)
# Error in data[[i]][rows, ] : subscript out of bounds
```










