---
title: "Continuous character models"
output:
  html_document:
    df_print: paged
---

Following along with http://phylometh.info/continuous.html
```{r, eval = TRUE}
library(ape) # utility functions
library(geiger) # utility functions
library(OUwie)
library(phytools)
library(tidyverse)
library(RCurl)
```

Get the tree and data
```{r, eval = TRUE, echo = TRUE}
sq_tree <- read.tree(text=getURL("https://raw.githubusercontent.com/lukejharmon/pcm/master/datafiles/squamate.phy"))

continuous.data <- read.csv(text=getURL("https://raw.githubusercontent.com/lukejharmon/pcm/master/datafiles/brandley_table.csv"), stringsAsFactors = FALSE)

rownames(continuous.data) <- gsub(" ", "_", continuous.data$Species)
hist(continuous.data$SVL)
# The data is right skewed and bounded by 0, I will log-transform it.
```

Clean the data to make sure everything matches
```{r, eval = TRUE}
CleanData <- function(phy, data) {
  treedata(phy, 
           continuous.data, 
           sort = FALSE, 
           warnings = TRUE)
}

CleanData(sq_tree, continuous.data)

continuous.data <- continuous.data[!(continuous.data$Species == "Gonatodes albogularis" |
                                      continuous.data$Species == "Lepidophyma flavimaculatum" |
                                      continuous.data$Species == "Trachyboa boulengeri"),]

CleanData(sq_tree, continuous.data)

# to tidy up the data I picked only one continuous column. Snout-vent length (SVL) in mm.
# the data need to be in a vector, too. Not a dataframe. Log transform it to make it more normal.
df.subset.continuous.data <- log(continuous.data["SVL"])
subset.continuous.data <- df.subset.continuous.data[,1]
names(subset.continuous.data) <- rownames(df.subset.continuous.data)
hist(subset.continuous.data)

print(ape::Ntip(sq_tree))
print(length(subset.continuous.data))
```

Visualize the data
```{r, eval = TRUE}
VisualizeData <- function(phy, data) {
  contMap(tree = phy,
          x = data)
}
VisualizeData(sq_tree, subset.continuous.data)
```


```{r, eval = TRUE}
plot.phylo(sq_tree)
```

Rate of evolution
```{r, eval = TRUE}
BM1 <- geiger::fitContinuous(sq_tree, subset.continuous.data, model = "BM")

print(paste("The rate of evolution is", BM1$opt$sigsq, "in units of", "log(mm^2) / time"))
```

Changing alpha changes how much stretch there is. How to determine alpha?
```{r, eval = TRUE}
OU1 <- fitContinuous(sq_tree, subset.continuous.data, model = "OU")
par(mfcol = c(1,2))
plot(sq_tree, show.tip.label = FALSE)
ou.tree <- rescale(sq_tree, model = "OU", .01)
plot(ou.tree, show.tip.label = FALSE)
```

The tree with the alpha transformation had longer tip lengths and shorter internal branches. 
Now compare the two models
```{r, eval = TRUE}
AIC.BM1 <- print(BM1$opt$aic)
AIC.OU1 <- print(OU1$opt$aic)
delta.AIC.BM1 <- if(AIC.BM1 < AIC.OU1) {
  (AIC.BM1 - AIC.BM1)
} else {
  (AIC.BM1 - AIC.OU1)
}
delta.AIC.OU1 <- if(AIC.OU1 < AIC.BM1) {
  (AIC.OU1 - AIC.OU1)
} else {
  (AIC.OU1 - AIC.BM1)
}

print(paste("The AIC of BM1 is", AIC.BM1))
print(paste("The AIC of OU1 is", AIC.OU1))
print(paste("The deltaAIC of BM1 is", delta.AIC.BM1))
print(paste("The deltaAIC of OU1 is", delta.AIC.OU1))
```

Assign regimes. Use an ancestral state estimation of a discrete trait. Use just one discrete character.
```{r, eval = TRUE}
one.discrete.char <- continuous.data["Morph"]
# make the discrete character a vector
one.discrete.char <- one.discrete.char[,1]
names(one.discrete.char) <- rownames(continuous.data)

# find the likelihood of each value (0, 1, 2) at the nodes
reconstruction.info <- ace(x = one.discrete.char, phy = sq_tree, type="discrete", method="ML", CI=TRUE)

# look across the liklihoods of the three options in the discrete data and pick the best one.
best.states <- colnames(reconstruction.info$lik.anc)[apply(reconstruction.info$lik.anc, 1, which.max)]

# the best.states didn't have the same length... but it's the value at the nodes, so it should have one less value than tips. 
# I can get the next section to work if I arbitrarily add a number to the best.states when I make the dataframe. Probably a bad idea. 
# df <- data.frame(species = names(subset.continuous.data), regime = c(best.states, "0"), continuous = subset.continuous.data)
#### df <- data.frame(species = names(subset.continuous.data), regime = best.states, continuous = subset.continuous.data)
# Error in data.frame(species = names(subset.continuous.data), regime = best.states, : arguments imply differing number of rows: 258, 257
```

Add the labels to the tree.
```{r, eval = TRUE}
# labeled.tree <- ace(one.discrete.char, sq_tree, model = "ER", type = "discrete")

# labeled.tree <- make.simmap(sq_tree, one.discrete.char, model = "ER")

# plot(labeled.tree)
# nodeBased.OUMV <- OUwie(phy = labeled.tree, 
#                         data = df, #I need to figure out how to make the regimes the same length as the dataframe :|
#                         model = "OUMV", 
#                         simmap.tree = TRUE, 
#                         diagn = FALSE)
# print(nodeBased.OUMV)
```


```{r, eval = TRUE}
# models <- c("BM1","BMS","OU1","OUM","OUMV","OUMA","OUMVA")
# results <- lapply(models, 
#                   OUwie, # how do I use lapply to run a function over a list?
#                   phy = sq_tree, 
#                   data = subset.continuous.data)
# 
# AICc.values<-sapply(results, "[[", "AICc")
# names(AICc.values)<-models
# AICc.values<-AICc.values-min(AICc.values)
# 
# 
# print(AICc.values) #The best model is the one with smallest AICc score
# 
# best<-results[[which.min(AICc.values)]] #store for later
# 
# print(best) #prints info on best model
```

