---
title: "The effect of dormancy on extinction probability"
output:
  html_document:
    df_print: paged
---

Based on:
Lennon, J. T., den Hollander, F., Wilke-Berenguer, M. & Blath, J. Principles of seed banks and the emergence of complexity from dormancy. Nature Communications 12, (2021).

Usually only external forces such as environmental conditions or resource availability are considered when looking at fluctuations and competition between species, but the ability to enter dormancy can have a significant impact in how successful a species is during competition. This model will model how extinction rates across phylogeny vary with dormancy.

###########################
-Taxa with dormancy grow slower than non-dormancy
-Taxa with dormancy outcompete non-dormancy
-Taxa without dormancy grow faster than dormancy
-Taxa without dormancy are outcompeted by dormancy
-If no competition, species without dormancy will have a higher speciation rate and lower extinction rate
-If competition present, species with dormancy will have a higher speciation rate and lower extinction rate

SpN = Not dormancy capable species
SpD = Dormancy capable species
Rep = Reproduction rate
Dea = Death rate
Ext = Extinction rate
Pop = Population size
Esc = Escape rate
SR = SpeciationRate
ER = ExtinctionRate
```{r}
# SpD_Rep <-  # > 0, < Sp1_Re
# SpD_Dea <-  # < Sp2_Re
# SpD_Esc <- 
# SpD_Pop <- ((Sp2_Rep - Sp2s_Dea) / (1 - Sp2_Esc))
#   
# SpN_Rep <-  # > 0, > Sp2_Rep
# SpN_Dea <-  # < Sp2_Re
# SpN_Pop <- (Sp1_Rep - Sp1_Dea)
#   
# SpD_SR <- 
# SpD_ER <- 
# SpN_SR <- 
# SpN_ER <- 
```

```{r}
library("hisse")
library("diversitree")
```

Use phylogeny and trait data from homework 8 to make and test the model.
```{r}
require("corHMM") # just like library(), but is for packages that are used within a function.
# corHMM extimates hidden rates underlying the evolution of a binary character. It's a hidden rates model. 
data(primates)
ls() # lists all data and values in the environment
print(primates)
require(phytools)
```
I want to have all possible trait combinations, so I made everything in the Hylobates genera have 1 for T1 (warning: made-up data!)
```{r}
primates$trait[which(grepl("Hylobates", primates$trait[,1])),2] <- 1
print(primates)
```

```{r}
trait1 <- primates$trait[,2]
names(trait1) <- primates$trait[,1]
primates$tree <- ape::multi2di(primates$tree)
plotSimmap(make.simmap(primates$tree, trait1), 
           pts = FALSE, 
           fsize = 0.8)
```

# I might not need to make a correlation matrix...!?
-Make a correlation matrix: compare "equal rates" to "all rates different"
-This will help fill in the transition rates in "pars" in the tree.musse part.
-I want to keep some dual transitions (1A -> 0B and 0B -> 1A) because competition may be introduced as a result of losing dormancy and competition might be removed as a result of gaining dormancy. 
```{r}
trans.rates = TransMatMaker.old(hidden.states=TRUE)
dormancy.trans.rates = ParDrop(trans.rates, c(3, 10))
allequal.dormancy.trans.rates = ParEqual(dormancy.trans.rates,
                                         c(1,2,
                                           1,3,
                                           1,4,
                                           1,5,
                                           1,6,
                                           1,7,
                                           1,8,
                                           1,9,
                                           1,10))
allequal.dormancy.trans.rates
```
Supply this transition matrix to the trans.rate= argument in hisse()

I need to make turnover and extinction rates for a HiSSE model. And also make the two column matrix or data frame, with the first column containing the species names and the second containing the binary character information.
```{r}
turnover.anc = c(1, 2, 3, 4)
eps.anc = c(1, 2, 3, 4)
phy.dat = data.frame(names(phy$tip.state), phy$tip.state)
phy.dat = phy.dat
```


```{r}
pp = hisse.old(phy = primates,
               data = phy.dat,
               f = c(1,1),
               hidden.states = TRUE,
               turnover.anc = turnover.anc,
               eps.anc = eps.anc,
               trans.rate = allequal.dormancy.trans.rates)
```
Error in UseMethod("Ntip") : no applicable method for 'Ntip' applied to an object of class "list"

```{r}
# # makes a rate matrix, can make a matrix without this code if needed.
# # rate matrix maker in the corHMM package doesn't exist anymore, so use ::: to find it :)
# rate.mat.er <- corHMM:::rate.mat.maker(rate.cat = 1,
#                               hrm = FALSE,
#                               ntraits = 2,
#                               nstates = 4,
#                               model = "ER") # equal rates model
# # find the optimized rates
# pp.er <- corHMM(phy = primates$tree,
#                 data = primates$trait[,c(1,2)],
#                 rate.cat = 1,
#                 rate.mat = rate.mat.er,
#                 node.states = "marginal")
# 
# 
# rate.mat.ard <- corHMM:::rate.mat.maker(rate.cat = 1,
#                                         hrm = FALSE,
#                                         ntraits = 2,
#                                         nstates = 4,
#                                         model = "ARD") # ARD = all rates different model
# # find the optimized rates
# pp.ard <- corHMM(phy = primates$tree,
#                  data = primates$trait[,c(1,2)],
#                  rate.cat = 1,
#                  rate.mat = rate.mat.ard,
#                  node.states = "marginal")
# 
# 
# if (pp.er$AICc < pp.ard$AICc) {
#   print(paste("The equal rates model AICc (", pp.er$AICc, ") is lower than the all rates different model AICc (", pp.ard$AICc, ")"))
# } else {
#   print(paste("The all rates different model AICc (", pp.ard$AICc, ") is lower than the equal rates  model AICc (", pp.er$AICc, ")"))
# }
```
```{r}
# print(pp.er)
```
If I can figure out the rate matrix above, I will continue this method below. 

I am going to try and model this like a tree-plus-trait model (HiSSE).
The trait I know is 0 = no_dormancy, 1 = dormancy.
The hidden trait is A = no_competition, B = competition.

SpN = Not dormancy capable species
SpD = Dormancy capable species
Rep = Reproduction rate
Dea = Death rate
Ext = Extinction rate
Pop = Population size
Esc = Escape rate
SR = SpeciationRate
ER = ExtinctionRate

```{r}
speciation.rates <- c(0.2, 0.1, 0.1, 0.2) #0A, 1A, 0B, 1B
# chose higher speciation rates for:
#   0A (no dormancy, no competition)
#   1B (dormancy, competition)
extinction.rates <- c(0.1, 0.1, 0.5, 0.1)
# chose higher extinction rate for:
#   0B (no dormancy, competition)
transition.rates <- c(0.01,0.01,0, 0.01, 0, 0.01, 0.01,0,0.01, 0,0.01,0.01)
pars <- c(speciation.rates, extinction.rates, transition.rates)
phy <- tree.musse(pars = pars, 
                  max.taxa = 50, 
                  x0 = 1, 
                  include.extinct = FALSE) 
# vector of pararmeters. Same order as unconstrained likelihood function returned by make.x for tree type x.
# maximum number of taxa in tree
# initial character state at root
```
